An examination of the accessibility implications of a pilot COVID-19 vaccination program in Hamilton, Ontario

### 2021-05-03 Accessibility-Vaccination-Sites-Hamilton/Response-to-Reviewers-v2.Rmd

---
title: "Response to Reviewers"
author: "Antonio Paez & Christopher D. Higgins"
date: "5/11/2020"
output: pdf_document
---

We would like to thank two anonymous reviewers for their comments and suggestions for revisions. In this letter we respond to those comments and explain how the paper was revised in response.

\newpage

# Reviewer 1

This article examines the accessibility implications of a pilot COVID-19 vaccination program for adults age 55+ (focusing on adults age 55–69) in Hamilton, Ontario. It makes a timely contribution to accessibility literature, generally, and to a small but growing body of scholarship more specifically related to accessing the COVID-19 vaccine. This article is certainly publishable, but would benefit from minor revisions. I present some suggestions and comments following that the authors might consider as they revise the piece:

>\textcolor{blue}{Thank you for your positive assessment of the paper and suggestions for improvements. We feel that the paper is better with the revisions suggested.}

• Please clarify from the outset of the article that you all are measuring “cost of travel” in travel time.

>\textcolor{blue}{We have done this, starting from the abstract.}

• The initial description of the pilot program is a bit unclear. By my understanding, 20 pharmacies in Hamilton were added as vaccination sites, and people age 55+ qualified to be vaccinated at these 20 pharmacies. People age 55–69 do not qualify at the other 325 locations mentioned, correct? I suggest you all clarify this. Also, the authors very briefly mention mobile vaccination pop-up clinics at the end of the article. Who’s eligible for these? Where are they? How many are there? I think the analysis in your article is scoped appropriately, but I do wonder whether the accessibility “gains” garnered by this pilot program matter too much in the grand scheme, particularly if so many other stationary/mobile clinics are available or are likely to become available to the 55+ age group (or even more age groups) soon. I suppose I’m asking that the authors clarify the importance of understanding and analyzing this pilot program in light of the greater vaccine rollout in the area.

>\textcolor{blue}{We are grateful for this comment. In trying to be concise we did not manage to be clear, and we appreciate the opportunity to clarify. In this respect, it is important to note that the City of Hamilton operated dedicated vaccination sites for people 70+ and mobile pop-up clinics for people 75+\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}; however, it did not have any province-approved pharmacies as part of the pharmacy vaccination program until April 1st, when we began this research. The 325 pharmacies mentioned were all in other cities in Ontario, the closest of which is Toronto. In this way, the pilot greatly expanded access to the vaccine in the City of Hamilton, however our analysis suggests that the selection of sites was inequitable, requiring lower income populations in the city center and rural populations to travel much longer to reach a vaccination site. We revised the introduction and hope that the revisions will provide sufficient clarity with respect to the program and its relevance.}

• The authors write, “As Yu et al. (2021) note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may lead to inequities in access.” I understand word count is limited, but I encourage the authors to provide more discussion here of why this “may lead to inequities in access” and who, exactly, is likely to reap the benefits and bear the costs of these inequities.

>\textcolor{blue}{We provide more discussion about this in the introduction, in effect previewing our hypotheses for the paper.}

• The travel time threshold/walking distance cutoff values seem rather large. Did you test or consider other, smaller values?

>\textcolor{blue}{The thresholds were used to calculate the travel time tables. Since we don't know the distribution of travel times a priori, a conservative approach to ensure that every parcel can reach at least one pharmacy is to use large thresholds when running the routing algorithm. After the travel time tables have been processed the thresholds are not relevant, because the analysis is based on travel time to the nearest pharmacy. As seen in Figure 4 of the original paper, the longest trip to the nearest pharmacy from any location is approximately 40 minutes (i.e., there are lots of superfluous origin-destination pairs in our time tables).}

• The expected travel time incorporating weighted sums given the modal split in a TAZ is interesting. Did you consider analyzing/presenting findings for different modes discretely, which might highlight intramode inequities more clearly? I’m mostly curious—I do think the authors’ approach using the weighted sums capturing, in theory, travel time by all modes is a valid and interesting approach.

>\textcolor{blue}{Thank you for this excellent suggestion. We now report the estimated travel times by mode with the twenty pharmacies and with the proposed additions. The additions reduce travel time for every mode, but the largest reductions are for transit and walking. At a time when mobility is limited to essential travel, it clearly is in the public interest to reduce the need to travel longer by any mode.}

• I find it more interpretable and impactful when you all present travel time results measured in minutes rather than in person-hours. I understand needing the person-hours results as a cumulative, individual-based measure to compare across income and region classes. However, as a reader, I do find the person-hours analysis to be harder to understand than the travel time analysis. I suggest the authors put some more thought towards explaining, particularly, the person-hours analysis results clearly.

>\textcolor{blue}{We agree. The cumulative values are easier to understand in hours, but at the individual level it is less intuitive to say "0.09 hours". We have adjusted this in the revised paper.}

• I like the idea of the comparison scenario, but wonder how the scenario sites were chosen. Could the three sites feasibly serve as vaccination sites? Were these locations “optimized” in any way to lead to greater reductions in travel time disparities? Were they randomly chosen? I think more is needed to clarify this scenario experiment/analysis for the reader.

>\textcolor{blue}{The three sites we use for the scenario are existing pharmacies and of chains used by the provine elsewhere for the pilot, so they are definitely feasible for the program. We did not use any systematic approach to select these sites, we just went with three reasonable sites. The fact that three sites chosen in a reasonable but not systematic way can lead to such important reductions is telling about the effort that the province put in the selection of sites for the program. There are certainly better ways of approaching this issue (for instance, by designing optimal accessibility landscapes, see Paez et al., 2013) but we feel that designing optimal landscapes is beyond the scope of a short findings paper, since the finding that we wish to report is the inequity in the roll-out of the program.}

• The article would benefit from proofreading for spelling and grammar errors, as well as some issues with general fluency and, especially, clarity of language. For instance, I suggest “do not serve well urban and rural residents,” be rephrased as, “do not serve urban and rural residents well.” Also, a (potentially important) typo in Methods: “Using the population aged 55 to 59 y.o.”—do you all mean 55 to 69 y.o.?

>\textcolor{blue}{We proofread the paper and hopefully the overall writing is improved. In the methods we did indeed mean 55 to 69 y.o.; this has been fixed in the revision.}

• The authors should note that the Findings guidelines state, “All articles have a maximum of 1000 words of text, exclusive of References. In addition, there is a maximum figure count of 3, and maximum table count of 3.” I am not sure of the article’s current word count, but I do notice the article includes 4 figures. The authors might consider combining Figures 1 and 2 by layering the vaccination site locations atop the population distribution maps (although, this might look busy). Figure 2 might, alternatively, be included as supplemental information.

>\textcolor{blue}{We tried layering the locations of the pharmacies on the population maps, but legibility is severely compromised. We are open to including some of the figures as supplemental information at the pleasure of the editor. Perhaps the editor is willing to trade one figure for one table? (we only have two tables in the paper and both are relatively compact).}

I am glad to have had the opportunity to review this interesting paper!

>\textcolor{blue}{Thank you again for you suggestions, and the opportunity to revise and improve the paper.}

\newpage

# Reviewer 2

This paper looks at a location allocation issue that has been studied before in geograohy for several decades, where to place new facilities ti reduce travel and increase access to heakthcare in a region. The study finds that rural and low income are poorly srved and big burden of travel falls on these groups and they offer the location of new sites for vaccinations to decrease such burden. The paper depnds ona major assumotion that those who commuted by mode X will commute by the same mode for vaccination, which is not true, if COVID has shiwn us something is new travel behaviour emerge and peopel are switching modes so this assumption need to be highlighted as a major limitation. it could have been better to see a map with the old sites alone, with new sites added and have the background as the collective travel time this will enable the reader to visualize the impact spatially.

>>\textcolor{blue}{Thank you for these comments. Both authors are familiar with the literature on accessibility to healthcare, having contributed to it as recently as 2019. The findings that we wish to report here concern the specific situation of the vaccination pilot in Ontario. This is a policy-relevant issue for which there is no empirical information available besides this paper. We agree that a limitation of the analysis is that we assume all demographic/travel behavior patterns are stable since 2016. This assumption was already highlighted in the original paper (and aren't all assumptions limitations by definition?); in the revision we  redouble our effort to highlight this issue more explicitly as a limitation. We are aware that travel behavior for routine activities has changed during the pandemic (the first author has researched this in, for example, Paez, 2020, and Jamal and Paez, 2020, both papers published in Findings). But travel to get an essential vaccine is not necessarily a routine, quotidiane sort of travel behavior. We do not really know (at least for the region under study) how behavior for non-routine travel may have changed during the pandemic since we know relatively little about non-routine travel to begin with. In principle, it seems possible to do scenario analysis, but for this we would need to make up numbers regarding modal shares during the pandemic, and such an endeavor is not without problems either - besides probably requiring a substantially longer paper. We plotted the locations of the pharmacies in the travel time maps for ease of visualization.}

### 2021-05-03 Accessibility-Vaccination-Sites-Hamilton/Accessibility-Vaccination-Sites-Hamilton.Rmd
First revisions after reviewer reports from Findings

---
title: "An examination of the accessibility implications of a pilot COVID-19 vaccination program in Hamilton, Ontario"
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: University of Toronto Scarborough
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada
  - code: University of Toronto Scarborough
    address: Department of Geography & Planning, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON M1C1A4
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  The Government of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines. The accessibility of this program raises questions about the cost of travel and the distribution of the cost among the population. In our examination of the City of Hamilton we find that selected sites do not serve well rural and urban residents and that the associated cost of travel is expected to be disproportionally borne by lower income populations. Modest additions to the list of pilot sites in the city can substantially alleviate this inequity.

@@

@@

```{r load-packages, include=FALSE}
#Load the packages needed to read and work with the data
library(grid)
library(gridExtra)
library(kableExtra)
library(png)
library(sf)
library(tidyverse)
library(units)
library(vaccHamilton) # Data package
```

@@

```{r table-comparison, echo=FALSE}
table_comparison <- cbind(person_hours_baseline %>%
        dplyr::select(group, total_population, total_weighted_person_hours, ratio), 
      person_hours %>%
        dplyr::select(total_weighted_person_hours, ratio))
```

```{r comparison-travel-time-by-mode, echo=FALSE}
# Comparison of weighted travel times by mode between pilot and scenario
min_ttm_mode_baseline %>%
  group_by(Mode) %>%
  summarize(pht_mode = sum(Population * weighted_travel_time/60),
            .groups = "drop")

min_ttm_mode %>%
  group_by(Mode) %>%
  summarize(pht_mode = sum(Population * weighted_travel_time/60),
            .groups = "drop")
```

@@

Along with the provision of health care facilities to treat severe cases of COVID-19 [@Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs. The Government of Ontario, in Canada, announced on April 1st 2021 the expansion of a pilot program to offer vaccines in pharmacies in the City of Hamilton\footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}. This program is in addition to dedicated vaccination centers for people aged 70+. Twenty pharmacies in Hamilton were added to an earlier list of 325 locations in other cities across the province, and the program was extended to people aged 55 years old and over.

Critics were swift to point out that the list of pharmacies approved for Hamilton by the province were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}. Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton. The issue is less clear-cut when we consider that Hamilton's older population skews suburban (see Figure \ref{fig:population-map}). Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and the young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural. Nevertheless, the selection of sites by the province raises some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}. As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may lead to inequities in access.

@@

- What is the estimated cost of travel to reach the vaccination sites, assuming that every person requires a vaccine?
- What is the distribution of this cost across the population of the city?
- How does the cost and its distribution change with the addition of candidate sites in urban Hamilton?

We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot.

@@

Using the population aged 55 to 59 y.o. we first calculate the average number of people per dwelling and assign them proportionally to the dwellings by parcel. Second, median incomes and modal splits across walking, transit, and car are joined to the parcels. Although our data are from 2016 (the most recent available), we assume they are largely representative of current spatial and demographic trends in the city. Third, we use the `r5r` [@Pereira2021r5r] package to calculate the travel time from each parcel to all pharmacies by three modes using a cutoff value of 180 min and a maximum walking distance of 10,000 m. Once we obtained travel time tables with population, proportion of trips by mode, and income information, we calculated the expected travel time $ett$ from each parcel $i$ to a pharmacy $j$ as follows:
$$
ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
$$
\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the vector of travel times from parcel $i$ to the pharmacies. The expected travel time is thus the weighted sum of travel times to the nearest pharmacy, with the weights given by the expected modal split in the TAZ.

The expected travel time $i$ was multiplied by the assigned population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) as follows:
$$
PHT_i = P_i\cdot ett_i
$$

@@

The top panel of Figure \ref{fig:maps-baseline} shows the average expected use-weighted multi-modal travel times by TAZ in Hamilton. It is apparent that travel times tend to be lower in much of suburban Hamilton and higher in the urban core and some rural parts of the city, particularly to the west. This is unsurprising, given the higher probability of travel by car and the predominantly suburban character of the vaccination sites. However, even accounting for the distribution of population, this leads to large disparities in the number of person-hours of travel across the city, with a concentration of the burden of travel in the urban core and the rural west (see bottom panel of Figure \ref{fig:maps-baseline}).

@@

For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot. We repeat the analysis, but include the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}. The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}. While all income groups benefit from the addition of these three sites with shorter mean trip durations, the most remarkable difference is the large reduction in the disparities between residents in DAs with lower levels of income. The top-right panel of Figure \ref{fig:results} shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads. Unsurprisingly, the addition of three urban vaccination sites does not have a large impact for rural residents.

The results indicate that the locations chosen by the province for the pilot vaccination program do not serve well urban or rural residents of the city, and there are some important questions regarding equity of access to the program, with a disproportionate burden in the cost of travel falling on lower income urban populations and rural populations. Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation. On the other hand, because there are not many candidate locations in rural parts of the city, increasing access for residents there likely necessitates an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.

```{r figure-maps-baseline, echo=FALSE, warning=FALSE, fig.height=7, fig.align="center", fig.cap="\\label{fig:maps-baseline}Average expected travel time by TAZ (in minutes) and total person-hours of travel by TAZ."}
grid.arrange(ttm_map_baseline +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             pht_map_baseline +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             nrow = 2)
```

```{r save-figure-maps-baseline, include=FALSE}
jpeg(filename = "images/baseline-maps.jpeg")
grid.arrange(ttm_map_baseline +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             pht_map_baseline +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             nrow = 2)
dev.off()

@@

```{r table-results, echo=FALSE}
table_comparison  %>%
  kable("latex",
        booktabs = TRUE,
        col.names = c("Group", 
                      "Population", 
                      "Total PHT", 
                      "Hours per person", 
                      "Total PHT",
                      "Hours per person"),
        digits = 3,
        align = c("lccccc"),
        caption = "\\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only, and scenario with three urban locations added") %>% 
  add_header_above(c(" " = 2, "Pilot Program" = 2, "Scenario" = 2)) %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  pack_rows("Income Quintile", 1, 5) %>%
  pack_rows("Region", 6, 8) %>%
  footnote(general = "The population totals differ due to small differences in the classification of the regions")
```

>>>>

---
title: "An examination of the accessibility implications of a pilot COVID-19 vaccination program in Hamilton, Ontario"
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Christopher D. Higgins
    email: cd.higgins@utoronto.ca
    affiliation: University of Toronto Scarborough
address:
  - code: McMaster University
    address: School of Earth, Environment and Society, McMaster University, Hamilton, ON, L8S 4K1, Canada
  - code: University of Toronto Scarborough
    address: Department of Geography & Planning, University of Toronto Scarborough, 1265 Military Trail, Toronto, ON M1C1A4
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  The Government of Ontario in Canada announced the pilot for a new vaccination program, with designated pharmacies across the province now able to offer COVID-19 vaccines. The accessibility of this program raises questions about travel times to vaccination sites and the distribution of these times among the population. In our examination of the City of Hamilton we find that selected sites do not serve rural and urban residents well; particularly, the associated cost of travel (in terms of travel time) is expected to be disproportionally borne by lower income urban populations and rural residents. Modest additions to the list of pilot sites in the city can substantially alleviate this inequity.

@@

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

@@

```{r load-packages, include=FALSE, cache=FALSE}
#Load the packages needed to read and work with the data
library(grid)
library(gridExtra)
library(kableExtra)
library(png)
library(sf)
library(tidyverse)
library(units)
library(vaccHamilton) # Data package
```

@@

```{r table-comparison, echo=FALSE}
table_comparison <- cbind(person_hours_baseline %>%
        dplyr::select(group, total_population, total_weighted_person_hours, ratio), 
      person_hours %>%
        dplyr::select(total_weighted_person_hours, ratio))

colnames(table_comparison) <- c("group", "total_population", "total_weighted_person_hours_baseline", "ratio_baseline",
                                "total_weighted_person_hours", "ratio")

table_comparison <- table_comparison %>%
  mutate(total_weighted_person_hours_baseline = round(total_weighted_person_hours_baseline, 2),
         ratio_baseline = set_units(ratio_baseline, h) %>% 
           set_units(min) %>% 
           drop_units(),
         ratio_baseline = round(ratio_baseline, 2),
         total_weighted_person_hours = round(total_weighted_person_hours, 2),
         ratio = set_units(ratio, h) %>% 
           set_units(min) %>% 
           drop_units(),
         ratio = round(ratio, 2))
```

```{r comparison-travel-time-by-mode, echo=FALSE}
# Comparison of weighted travel times by mode between pilot and scenario
comparison_ttm_mode <- cbind(min_ttm_mode_baseline %>%
  group_by(Mode) %>%
  summarize(pht_mode_baseline = sum(Population * weighted_travel_time/60),
            .groups = "drop"),
  min_ttm_mode %>%
  group_by(Mode) %>%
  summarize(pht_mode = sum(Population * weighted_travel_time/60),
            .groups = "drop") %>%
    select(pht_mode)) %>% 
  mutate(diff = paste0(round((pht_mode - pht_mode_baseline)/pht_mode_baseline * 100, 2), "%", sep = ""))
```

@@

Along with the provision of health care facilities to treat severe cases of COVID-19 [e.g., @Ghorbanzadeh2021spatial; @Pereira2021geographic], another front in the fight against the pandemic is the rolling out of vaccination programs. The Government of Ontario, in Canada, implemented in 2021 a pilot program to offer vaccines in pharmacies for people 60 years and older. In the first incarnation of this program, the province approved 325 pharmacies in Toronto, Windsor-Essex, and Kingston (the latter including the region of Frontenac-Lennox and Addington). The City of Hamilton, the third most populous urban center in the province, operated dedicated vaccination centers for people aged 70+ and mobile pop-up clinics for people aged 75+; however, it did not have any approved pharmacies under the pilot for other populations until April 1st 2021, with the announcement of 20 pharmacies in the city \footnote{\url{https://www.cbc.ca/news/canada/hamilton/astrazeneca-vaccine-hamilton-1.5972704}}. At the same time that the expansion of the pilot was announced, the province extended eligibility for this program to people 55 years and older.

Critics were swift to point out that the list of pharmacies approved for Hamilton by the province were mostly located in lower density parts of the city that are not well serviced by transit and are difficult to reach by foot\footnote{See inter alia: \url{https://twitter.com/RyanMcGreal/status/1378027149790224386?s=20} and \url{https://twitter.com/NrinderWard3/status/1378679195514060801?s=20}.}. Indeed, as seen in Figure \ref{fig:pharmacies-and-regions}, a vast majority of the pharmacies are in suburban Hamilton. Whether this is an issue is less clear-cut when we consider that Hamilton's older population skews suburban (see Figure \ref{fig:population-map}). Given the target demographic for the program, it is possible that suburban sites could be convenient for mature adults and the young old: the population aged 55 to 69 in Hamilton is approximately `r prettyNum(round((table_comparison[7,2]/1000), 2) * 1000, big.mark = ",")` suburban, `r prettyNum(round((table_comparison[8,2]/1000), 2) * 1000, big.mark = ",")` urban, and only `r prettyNum(round((table_comparison[6,2]/1000), 2) * 1000, big.mark = ",")` rural. Nevertheless, the selection of sites by the province raised some important questions\footnote{The decision-making process to select these sites appears to have been opaque, and the Mayor of the city was caught flat footed by the announcement; see: \url{https://twitter.com/FredEisenberger/status/1378350123114242053?s=20}}. As @Yu2021sustained note, good geographical coverage is a key element for a successful vaccination campaign; at the same time, siting vaccinations sites in car-oriented locations may lead to inequities in access by requiring people who do not own cars or who typically travel by other modes (mostly lower income urban residents) to spend more of their time reaching an approved pharmacy.

@@

- What is the estimated travel time needed to reach the nearest approved pharmacy, assuming that every person requires a vaccine?
- What is the distribution of travel times across the population of the city?
- How does the cost of time needed for travel and its distribution change with the addition of candidate vaccination sites in urban Hamilton?

We concentrate on the 55 to 69 years old population segment because the older 70+ group have access to other dedicated facilities besides those in the provincial pilot (including pop-up clinics for people 75+).

@@

Using the population aged 55 to 69 y.o. we first calculate the average number of people per dwelling and assign them proportionally to the dwellings by parcel. Second, median household incomes and modal splits across walking, transit, and car are joined to the parcels. Our data are from 2016 (the most recent available), but we assume they are largely representative of spatial, mobility, and demographic trends in the city. In the absence of travel behavior information during the pandemic we suggest this is an inevitable limitation of the analysis. Third, we use the `r5r` [@Pereira2021r5r] package to calculate the travel time from each parcel to all pharmacies by three modes using a cutoff value of 180 min and a maximum walking distance of 10,000 m. These thresholds are chosen to maximize the coverage of the travel time tables, but the analysis, as explained next, is based on travel to the nearest approved pharmacy. Once we obtained travel time tables with population, proportion of trips by mode, and income information, we calculated the expected travel time $ett$ from each parcel $i$ to a pharmacy $j$ as follows:
$$
ett_i = p^c_i \min(tt^c_{ij}) + p^t_i \min(tt^t_{ij}) + p^w_i \min(tt^w_{ij})
$$
\noindent where $p^k_i$ is the proportion of trips by mode $k$ in the TAZ of parcel $i$, and $tt^k_{ij}$ is the vector of travel times from parcel $i$ to the pharmacies. The expected travel time is thus the weighted sum of travel times to the _nearest_ pharmacy, with the weights given by the expected modal split in the TAZ.

The expected travel time $i$ was multiplied by the assigned population in parcel $i$ to obtain a measure of person-hours of travel ($PHT$) that gives a cumulative cost (in travel time) for a population group. This is as follows:
$$
PHT_i = P_i\cdot ett_i
$$

@@

The top panel of Figure \ref{fig:maps-baseline} shows the average multi-modal travel times (weighted by modal shares) by TAZ in Hamilton. It is apparent that travel times tend to be lower in much of suburban Hamilton and higher in the urban core and some rural parts of the city, particularly to the west. This is unsurprising, given the higher probability of travel by car and the predominantly suburban character of the vaccination sites. However, even accounting for the distribution of population, this leads to large disparities in the number of person-hours of travel across the city, with a concentration of the burden of travel in the urban core and the rural west (see bottom panel of Figure \ref{fig:maps-baseline}).

@@

For comparison purposes we consider a scenario with some modest additions to the list of pharmacies in the provincial pilot. For the scenario we choose three plausible sites (i.e., pharmacies that belong to chains already used by the province for this program). We do not conduct a systematic search of candidate locations. We note that some papers propound the design optimal accessibility landscapes [e.g., @Horner2008optimal; @Paez2013exploring], but such an endeavor is beyond the scope of a short findings paper. Instead, we repeat the analysis after including the three urban sites shown in white circles in Figure \ref{fig:pharmacies-and-regions}.

The results of this scenario appear in the last two columns of Table \ref{tab:distribution-results} and the two right panels of Figure \ref{fig:results}. While all income groups benefit from the addition of these three sites with shorter mean trip durations, the most remarkable difference is the large reduction in the disparities between residents in DAs with lower levels of income. The top-right panel of Figure \ref{fig:results} shows that the distribution of expected travel time is now more in line for all income groups, even if the bottom two income quintiles still have somewhat wider spreads. Unsurprisingly, the addition of three urban vaccination sites does not have a large impact for rural residents. The scenario also suggests that important reductions in the need for travel can be achieved. While this is true for all three modes of transportation (see Table \ref{tab:distribution-results-by-mode}), the reductions are more substantial for travel by transit and walking. At a time that non-pharmaceutical interventions call for reductions in mobility, it is clearly in the public interest to reduce the need to travel even for this essential purpose.

The results indicate that the locations chosen by the province for the pilot vaccination program do not serve urban or rural residents of the city well, and there are some important questions regarding equity of access to the program. As shown, a disproportionate burden in the cost of travel (in terms of travel time) fall on lower income urban populations and rural populations. Selection of three sensible urban locations does much to alleviate disparities in the burden of transportation - and the fact that this is so despite the lack of a systematic search for optimal locations does not speak well of the effort put by the province in the choice of vaccination sites. The situation is different for rural populations, mainly because there are not many candidate locations in rural parts of the city; in this case, it is likely that increasing access for residents there will necessitate an expansion of the existing mobile vaccination pop-up clinic program\footnote{\url{https://www.hamilton.ca/government-information/news-centre/news-releases/hamiltons-covid-19-vaccination-program-expansion-1}}.

```{r figure-maps-baseline, echo=FALSE, warning=FALSE, fig.height=7, fig.align="center", fig.cap="\\label{fig:maps-baseline}Average expected travel time by TAZ (in minutes) and total person-hours of travel by TAZ."}
grid.arrange(ttm_map_baseline +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Pilot"),
                       shape = 17,
                       size = 1,
                       color = "black",
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 16,
                       color = "white",
                       size = 1,
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 1,
                       color = "black",
                       size = 1,
                       show.legend = FALSE) +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             pht_map_baseline  +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Pilot"),
                       shape = 17,
                       size = 1,
                       color = "black",
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 16,
                       color = "white",
                       size = 1,
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 1,
                       color = "black",
                       size = 1,
                       show.legend = FALSE) +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             nrow = 2)
```

```{r save-figure-maps-baseline, include=FALSE}
jpeg(filename = "images/baseline-maps.jpeg")
grid.arrange(ttm_map_baseline +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Pilot"),
                       shape = 17,
                       size = 1,
                       color = "black",
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 16,
                       color = "white",
                       size = 1,
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 1,
                       color = "black",
                       size = 1,
                       show.legend = FALSE) +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             pht_map_baseline  +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Pilot"),
                       shape = 17,
                       size = 1,
                       color = "black",
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 16,
                       color = "white",
                       size = 1,
                       show.legend = FALSE) +
               geom_sf(data = pharmacy_locations %>%
                         dplyr::filter(Type == "Candidate"),
                       shape = 1,
                       color = "black",
                       size = 1,
                       show.legend = FALSE) +
               theme(legend.title = element_text(size = 8),
                     legend.text = element_text(size = 7), 
                     axis.text = element_text(size = 7)),
             nrow = 2)
dev.off()

@@

```{r table-results, echo=FALSE}
table_comparison  %>%
  kable("latex",
        booktabs = TRUE,
        col.names = c("Group", 
                      "Population", 
                      "Total PHT", 
                      "Minutes per person", 
                      "Total PHT",
                      "Minutes per person"),
        digits = 3,
        align = c("lccccc"),
        caption = "\\label{tab:distribution-results}Distribution of person-hours of travel (PHT) by median total household income and region: pilot locations only (baseline), and scenario with three urban locations added") %>% 
  add_header_above(c(" " = 2, "Pilot Program" = 2, "Scenario" = 2)) %>%
  kable_styling(latex_options = c("striped", "scale_down")) %>%
  pack_rows("Income Quintile", 1, 5) %>%
  pack_rows("Region", 6, 8) %>%
  footnote(general = "The population totals differ due to small differences in the classification of the regions")
```

```{r table-results-by-mode, echo=FALSE}
comparison_ttm_mode %>%
  kable("latex",
        booktabs = TRUE,
        col.names = c("Mode", 
                      "Baseline PHT", 
                      "Scenario PHT", 
                      "Reduction in PHT"),
        digits = 3,
        align = c("lccc"),
        caption = "\\label{tab:distribution-results-by-mode}Distribution of person-hours of travel (PHT) by mode of transportation: pilot locations only (baseline) and scenario with three urban locations added")  %>%
  kable_styling(latex_options = c("striped"))
```
